Identifying optimized decision criteria and experimental designs by simulating preclinical experiments in silico

### 2020-09-23 Manuscript/Manuscript_intro_methods_results.Rmd
manuscript updated

When planning a replication study, the sample size has to be determined *a priori*. This is usually done using an effect size estimate from prior research. This can be taken from the literature or a pilot study. In both cases, the effect size estimate used is most likely inflated due to reasons outlined above. Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects. This is especially true if effects are small. When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists. An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.

@@

<!-- \ \ \ \ This simulation-based investigation explores different approaches to design and conduct experiments in preclinical settings. We explore preclinical research trajectories and determine efficiency and feasibility qua positive predictive value (PPV) and number of animals tested. These criteria ensure that while the number of animals is kept at a minimum, low sample sizes do not compromise the conclusiveness of the study. -->

@@

Drawing from empirical effect size distributions, enables the direct comparison between pre-study odds and the positive predictive value across the trajectory. If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.

@@

## Trajectories and outcome variables

@@

In the results section, we compare only the two trajectories that are the extremes of the spectrum of trajectories we have modeled: First, conventional significance threshold -- standard sample size calculation using initial
effect size estimate, second, SESOI within CI around initial effect size estimate -- standard sample size calculation using SESOI. In the following, we refer to the first trajectory as T1 and to the second as T2. We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).

T1 constitutes our reference, as we consider it to be closest to current practice. T1 and T2 are compared regarding the number of experiments proceeding to the replication stage, number of animals needed in the replication, and positive predictive value across the trajectory. Secondary outcomes are the false positive rate, false negative rate, and effect size precision.

@@

\ \ \ \ Given the pre-study odds of finding an effect of a certain size (in our case 0.5 and 1.0 as SESOI), this outcome is an indication of how strict or lenient the filters are we use to select effects sizes that we deem relevant for further examination (i.e. to confirm them in a within-lab replication study). We employed two different decision criteria to determine when to continue to the replication stage.

@@

\ \ \ \ The positive predictive value (PPV) of a study is the post-study probability that a positive finding which is based on statistical significance reflects a true effect [@ioannidis2005most]. To calculate the positive predictive value, one (ideally) needs the pre-study odds (prevalence), as well as the sensitivity and specificity of the test (measurement).

@@

\ \ \ \ The false positive rate (FPR) is the proportion of tests that detected a positive result when there was no true effect present. In our simulation, a false positive correponds to a significant finding given an underlying true effect $<$ 0.5 or 1.0, respectively.

@@

\ \ \ \ The false negative rate (FNR) is the rpoportion of tests that missed a true positive relationship. In our simulation, this correponds to non-significant results given that the underlying true effect was $\geq$ 0.5 or 1.0, respectively.

>>>>

When planning a replication study, the sample size has to be determined *a priori*. This is usually done using an effect size estimate from prior research. This can be taken from the literature or a pilot study. In both cases, the effect size estimate used is most likely inflated due to reasons outlined above. Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects. This is especially true if true underlying effects are small. When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists. An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.

@@

@@

@@

### Replication stage

\ \ \ \ For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate) a replication study was performed. The number of replication studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0). A replication study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size. For a replication be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05)

## Trajectories

@@

In the results section, we compare only the two trajectories that are the extremes of the spectrum of trajectories we have modeled: The first trajectory uses the conventional significance threshold as decision criterion and employs a standard sample size calculation using initial effect size estimate. The second trajectory applies a SESOI as decision criterion to move from the exploratory to the replication stage, and to calculate the sample size for the replication. In the following, we refer to the first trajectory as T1 and to the second as T2. We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).

T1 constitutes our reference, as we consider it to be closest to current practice. T1 and T2 are compared regarding the number of experiments proceeding to the replication stage, number of animals needed in the replication, and positive predictive value across the trajectory. Secondary outcomes are the false positive rate, false negative rate, and effect size precision. Outcome variables are outlined in more detail in the following section.

## Outcome variables

@@

\ \ \ \ This outcome serves as an indication of how strict or lenient the filters are that we used to select effect sizes that we deem relevant for further examination (i.e. to confirm them in a within-lab replication study).

@@

\ \ \ \ The positive predictive value (PPV) of a study is the post-study probability that a positive finding which is based on statistical significance reflects a true effect [@ioannidis2005most]. To calculate the positive predictive value, one (ideally) needs the pre-study odds (prevalence), as well as the sensitivity and specificity of the test (measurement). Given that we drew effect sizes from empirical effect size distributions, we know the pre-study odds of an effect of 0.5 and 1.0. Drawing from empirical effect size distributions, enables the direct comparison between pre-study odds and the positive predictive value across the trajectory. If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.

@@

\ \ \ \ The false positive rate (FPR) is the proportion of tests that detected a positive result when there was no true effect present. In our simulation, a false positive correponds to a significant finding given an underlying true effect smaller than 0.5 or 1.0, respectively.

@@

\ \ \ \ The false negative rate (FNR) is the proportion of tests that missed a true positive relationship. In our simulation, this correponds to non-significant results given that the underlying true effect was greater or equal to 0.5 or 1.0, respectively.
